---
header-includes:
- \usepackage{graphicx,latexsym}
- \usepackage{amssymb,amsthm,amsmath}
- \usepackage{longtable,booktabs,setspace}
output: word_document
---
# Analysis {.unnumbered}
  \setcounter{chapter}{5}
	\setcounter{section}{0}
  \doublespacing

Before examining the question of why an individual would use urgent care as their primary care physician, an exploratory cluster analysis of urgent care visits between 2008 and 2012 was performed. The goal here was not predictive; rather, for the purposes of this analysis, clustering was used as a form of reproduction--observed trends were used to generate research hypotheses which account for the observed groupings. Following this, two logistic regressions were performed on the characteristics which were identified through the clustering, with the goal of obtaining a model which addresses whether or not an individual would a) choose to go to urgent care and b) use urgent care as their primary care physician. 

## Identifying Typologies with Clustering{.unnumbered}

#### Differences in Demographic and Behavioral Factors of Urgent Care Seekers{.unnumbered}

In order to better examine the hypotheses about individuals choosing to go to urgent care, I first identified which characteristics or behaviors most distinguished sub-groups of urgent care patients in the cluster analysis. Testing the cluster analysis on critical demographic characteristics and visit behaviors that were discussed in the theory previously revealed clusters which qualitatively fit all three of the theoretical case profiles. The cluster analysis makes immediately clear that there are distinct homogeneous groups within the selection of patients who utilize urgent care. Figure 1 on the following page illustrates the observed sub-groups.

\newpage


```{r cluster_table, results = "asis", echo=FALSE}

reedtemplates::label(path = "figures/clusters.png", caption = "Identified patient typologies of urgent care centers visits (2008-2012).", 
      label = "clus1", 
      type = "figure", 
      scale = .45,
      angle = 90,
      options = "h!")
```

\newpage

Beginning with demographic trends, the characteristics which were *not* identified as distinguishable in the sample by the clustering (and were thus not included in Table 1.) offer just as much, if not more insight as those that did result in fractioning. Race in particular was immediately noticeable for its lack of variety: of the initial clusters, all were predominantly or exclusively white, an unsurprising trend when one looks at the summary statistics (of the 3863 urgent care visits included in the data, 87% of those were white individuals). This is substantively interesting for several reasons, mainly in that race is a known correlate with many other socioeconomic indicators, and urgent care appears to be largely utilized by white patients.

Similarly, no significant clusters of patients going on weekends were identified, nor were any of the major clusters' visits injury related. These observations together seem to counter the typical view of urgent care, and coupled with the fact that only 25% of the urgent care visits were coded as new visits, there are obvious indications that urgent care is being used in ways other than those proposed by hypothesis 1.

Since the data set lacked individual level socioeconomic data, we can examine the socioeconomic indicators for the patients’ ZIP codes which were included as demographic parameters proxies. Unsurprisingly, these appeared to cluster and correlate with both race and income: for both males and females, there were consistent clusters of White, 25-44-year-old patients with private insurance, and both of these clusters were in the highest quartile of percent population with Bachelor’s degrees, the second lowest percent population under the poverty line (5-10%), and both visited clinics coded as being located in a “Large central metro” (Table 1.). 

In fact, if one predominant trend becomes clear from the demographic clustering, it is that urgent care centers are almost entirely an urban affair. Of the clusters found, all but one consisted of visits located in medium to large metros, indicating that while other variation between patients exists, most urgent care consumers are located in larger cities. And if its singularity wasn’t enough, the cluster of visits which did occur in a rural setting also differed from most other clusters in the other parameters as well. As can be seen in the first row of Table 1., the rural cluster of visits had one of the highest levels of localized poverty, one of the lowest levels of educational attainment, and was one of only two clusters where a significant number of visits were paid with medicare.

Also informative are the clusters’ most occurring payment types, which complicate some of the theoretical notions of urgent care described in Chapter 1. While the largest clusters were consistently comprised of private insurance payers, Medicare without a doubt plays a role in getting people to urgent care centers. Medicare was listed as the primary payment method around 18 % of the time, a large number when one considers the consumerist theories proposed in chapter one. These clusters are crucial to understanding the decision process behind choosing a health care provider, while simultaneously raising the question of the relationship between the surprising number of elderly patients on Medicare seeking treatment at an urgent care centers.

The demographic characteristics of urgent care seekers were also clustered with behavioral, visit-level parameters in a second wave of cluster analysis, in order to to examine *how* urgent care is utilized by different groups. By far the largest and most homogeneous clusters isolated consist of white, privately insured patients in medium to large urban environments. These trends match those found in the demographic clusters, and we can learn even more about this group by examining the cluster segmentation that occurred across their behavioral parameters in the second wave. Of the two clusters which consist almost entirely of white, 25-44-year-old men and women, both were almost entirely classed as “established patients”, with at least one past visit. Also theoretically interesting, neither cluster had observations whose reason for visiting was “injury related” and neither had recorded visits on weekends. Clusters 5 and 6 in particular (Rows 5 and 6, Table 1), consisted of 25-64-year-old white patients with private insurance, an established history of visiting urgent care, and a reason for their current visit coded as a “chronic, routine problem”. 

In fact, only one cluster of visits in the final model were coded as new patients, though it is a sizable fraction of the whole.Ecxamining these two clusters reveals what I referred to earlier as the traditional patient characteristics ascribed to urgent care visits. These two cluster show what many would expect, both consisted of white, privately insured, new patients, and both clusters contain the only observations which occurred on a weekend or were coded as injury related.  These characteristics are in line with what many hold as the purpose of urgent care centers, but they remain a small fraction of the recorded number of visits. Completely contradicting this idea, the other seven of the nine clusters identified have high levels of established patients with chronic or routine problems, with ages ranging for 15 to over 75.
	
As for the Medicare question, if the clusters which consist of primarily elderly patients are examined, we can begin to understand a little better the relatively large percentage of what many would consider non-emergent care utilizers and their relationship to urgent care. Cluster 5 consists entirely of white patients over 74, whose visits were recorded as routine and who have a minimum of 3 past visits at the same center. Cluster three on the other hand has a slightly younger age range, and the majority of patients indicated that the urgent care center was *not* their primary care. Also distinguishing, cluster three is the only cluster with a majority of rural observations. When compared, it appears that there are two typologies of elderly urgent care seekers, the first of which may be using urgent in much the same ways as the younger and larger clusters of privately insured urbanites, while the second seems to rely on urgent care much more as an actually resource for urgent problems. 

In the end, the observed typologies appear to correspond relatively well with all three hypotheses presented, with the exception of a relatively large number of very young urgent care goers who did not list it as their primary care physician. The explanation for this is probably as simple as them having a pediatrician, and using urgent care as a quick supplement, which somewhat supports hypothesis 3. What the cluster analysis really does is inform the analysis about which variables were identified as important to urgent care seekers. Using these, we can now turn to a logistic regression in order to understand better how each parameter influences decisions regarding urgent care


## Logistic Regression Analysis {.unnumbered}

The initial exploratory analysis shows clear typologies of urgent care seekers which align with the hypotheses presented in the previous chapter, and the next phase of the analysis uses those trends and tests for significance using logistic regression analysis. Tables 5.1 and 5.2 show the results of the two logistic regression models. 

The first table is a model of the initial decision to go to urgent care. For the odds of choosing to go to urgent care over other means of ambulatory care, there is no significance for whether or not the visit is injury related, and whether or not it is a weekend. This supports the observations from the cluster analysis, and provides further evidence against the demand oriented hypothesis. Similarly, visits for routine and chronic issues have a higher chance of leading to an urgent care visit than problems less than 3 mos old. These trends both support Hypothesis 3: that urgent care centers are being used as a new mode of primary care.

For the second analysis, the outcome is no longer whether or not the visit was *at* urgent care, but whether or not an urgent care patient listed the urgent care center as their "Primary Health Provider". Interestingly, most of the significant variables' effects change direction when this is taken as the outcome. While high local income levels had a negative relationship with the decision to go to urgent care, it appears to very positively correlate with those who do use urgent care as their primary care. 

Many of the visit-specific variables that did not have a significant effect on the odds of choosing urgent care became significant when looking at urgent care as primary care. Unsurprisingly, whether or not the visit was injury related has a negative effect on the odds of the patient specifying that clinic as their primary care. 

There are some surprising trends in the secondary logistic model which deserve further study. Established patient, for example, have a significant negative correlation with the odds of a patient using urgent care as their primary care physician, as did number of past visits. A further look into the structure of the data reveals a plausible explanation: while 94% of patients who listed the urgent care center as their primary care physician were recorded as Established patients, this group as a whole only constitutes a third of the total number of visits to urgent care. In other words, there were two-thirds as many urgent care visits who were established patients, yet did *not* list the facility as their primary care office, and this trend is similar for number of past visits. 


```{r logit_regression, echo=FALSE, warning=FALSE, message=FALSE, cache=TRUE}
NAMCS.logit <- NAMCS.work
                        
# variable selection
logit.vars <- c("URGCARE","PRIMCARE","PCTPOVR", "HINCOMER", "PBAMORER", "URBANRUR", "RACER", "WEEKEND", "AGER", "SEX","PAYTYPER",
                "INJURY","SENBEFOR", "PASTVIS", "MAJOR")
logit.varset <- NAMCS.logit[logit.vars]
logit.varset <- na.omit(logit.varset)

# recoding
logit.varset$ZipHighIncome <- as.numeric(logit.varset$HINCOMER == "Quartile 4 ($52,388 or more)")
logit.varset$HighZipPov <- as.numeric(logit.varset$PCTPOVR == "Quartile 4 (20.00 percent or more)")
logit.varset$HighEduc<- as.numeric(logit.varset$PBAMORER == "Quartile 4 (31.69 percent or more)")
levels(logit.varset$URBANRUR) <- list(Urban=c("Large central metro", "Large fringe metro", "Medium metro"), 
                                      Rural=c("Micropolitan/noncore (nonmetro)", "Small metro"))
levels(logit.varset$AGER) <- list(YoungAdult=c("Under 15 years", "15-24 years"),
                                  WorkingAge=c("25-44 years", "45-64 years"),
                                  Retired=c("65-74 years", "75 years and over"))

levels(logit.varset$PAYTYPER) <- list(PrivateIns=c("Private insurance"),
                                  Medicaid=c("Medicaid"),
                                  Medicare=c("Medicare"),
                                  Other=c("Other", "No charge", "Unknown","Worker's compensation", "All sources of payment are blank", "Self-pay" ))

levels(logit.varset$MAJOR) <- list(NewProblem=c("New problem (less than 3 mos. onset)"),
                                      RoutineChronic=c("Pre-/Post-surgery", "Chronic problem, routine", "Chronic problem, flare-up"),
                                      Preventative=c("Preventive care"),
                                      Other=c("Blank", "No charge", "Pre-/Post-surgery"))

logit.varset$PASTVIS <- as.numeric(logit.varset$PASTVIS)


#urgentcareonly model
UC.varset <-subset(logit.varset, logit.varset$URGCARE == "Yes")

# make model matrices
dummiesFull.df <- as.data.frame(model.matrix(~., data=logit.varset)[,-1]) 
dummiesUC.df <- as.data.frame(model.matrix(~.,data = UC.varset)[,-1])


dummy.vars <- c("HighEduc", "HighZipPov", "ZipHighIncome", "MAJORPreventative", "MAJORRoutineChronic", "PASTVIS", "SENBEFORYes, established patient", "INJURYYes", "SEXMale", "AGERRetired", "AGERWorkingAge", "WEEKENDYES", "RACERWhite", "RACEROther", "URBANRURRural", "PRIMCAREYes", "URGCAREYes", "PAYTYPERMedicaid", "PAYTYPERMedicare", "PAYTYPEROther")
dummyFull.varset <- dummiesFull.df[dummy.vars]
dummyUC.varset <- dummiesUC.df[dummy.vars]

names(dummyFull.varset)[names(dummyFull.varset) == 'SENBEFORYes, established patient'] <- 'EstablishedPatient'
names(dummyUC.varset)[names(dummyUC.varset) == 'SENBEFORYes, established patient'] <- 'EstablishedPatient'

### summary stats ##

#summary(dummy.varset)

## LOGIT ##
#head(dummy.varset)
#summary(dummy.varset)
#sapply(dummy.varset, sd)
#xtabs(~ URGCAREYes + RACERWhite, data = dummy.varset)

logit.dem <- glm(URGCAREYes ~ INJURYYes + WEEKENDYES+ PAYTYPERMedicaid+PAYTYPERMedicare+PAYTYPEROther+EstablishedPatient + PASTVIS + MAJORPreventative + MAJORRoutineChronic, data = dummyFull.varset, family = binomial)
logit.beh <- glm(URGCAREYes ~ AGERWorkingAge + AGERRetired + SEXMale + RACERWhite + ZipHighIncome + URBANRURRural, data = dummyFull.varset, family = binomial)
logit.full <- glm(URGCAREYes ~ INJURYYes + WEEKENDYES+ PAYTYPERMedicaid+PAYTYPERMedicare+PAYTYPEROther+EstablishedPatient + PASTVIS + MAJORPreventative + MAJORRoutineChronic + AGERWorkingAge + AGERRetired + SEXMale + RACERWhite + ZipHighIncome + URBANRURRural, data = dummyFull.varset, family = binomial)


logit2.dem <- glm(PRIMCAREYes ~ INJURYYes + WEEKENDYES+ EstablishedPatient + PASTVIS + MAJORPreventative + MAJORRoutineChronic, data = dummyUC.varset, family = binomial)

logit2.beh <- glm(PRIMCAREYes ~ AGERWorkingAge + AGERRetired + SEXMale + RACERWhite + ZipHighIncome + URBANRURRural, data = dummyUC.varset, family = binomial)

logit2.full <- glm(PRIMCAREYes ~ INJURYYes + WEEKENDYES + PAYTYPERMedicaid + PAYTYPERMedicare + PAYTYPEROther + EstablishedPatient + PASTVIS + MAJORPreventative + MAJORRoutineChronic + AGERWorkingAge + AGERRetired + SEXMale + RACERWhite + ZipHighIncome + URBANRURRural, data = dummyUC.varset, family = binomial)

#summary(logit2.full)

#psudeor2 <- 1 - (logit.full$deviance / logit.full$null.deviance)

#psudeor2

#1−ResidualDevianceNullDeviance


```


\singlespacing
```{r logit_tables, warning=FALSE, message=FALSE  ,echo=FALSE ,results="asis"}

library(texreg)
texreg(list(logit.beh, logit.dem, logit.full), 
       caption = "Logistic model of the decision to go to urgent care.", 
       fontsize = "small", 
       custom.coef.names = c("Intercept", "Injury Related Visit", "Weekend Visit", "Medicaid" , "Medicare" , "SelfPay", "Established Patient", "Number Past Visits", "Visit Reason: Preventative", "Visit Reason: Routine/Chronic", "Age: Working", "Age: Retired", "Male", "Race: White", "High ZIP Income", "Rural"))

texreg(list(logit2.beh, logit2.dem, logit2.full), 
       caption = "Logistic model of the decision to use urgent care as primary care", 
       fontsize = "small",
       custom.coef.names = c("Intercept", "Injury Related Visit", "Weekend Visit", "Medicaid" , "Medicare" , "SelfPay" , "Established Patient", "Number Past Visits", "Visit Reason: Preventative", "Visit Reason: Routine/Chronic", "Age: Working", "Age: Retired", "Male", "Race: White", "High ZIP Income", "Rural"))


```


```{r diagnostics, warning=FALSE, message=FALSE  ,echo=FALSE}

#library(ResourceSelection)
#h1<- hoslem.test(dummiesUC.df$PRIMCAREYes, fitted(logit2.full, g =10))
#h1

#library(tree)
#urgentCareOnly.df$PRIMCAREYES <- ifelse(urgentCareOnly.df$PRIMCARE == "Yes",1,0)
#t1 <- tree(PRIMCAREYES ~. - PRIMCARE, data = urgentCareOnly.df, split = "gini")
#plot(t1)
#text(t1, pretty = 0)
#
#naCols <- vector(length=412)
#for (i in 1:412){ naCols[i] <- sum(is.na(urgentCareOnly.df[,i]))}
#subset <- urgentCareOnly.df[,which(naCols < 5)]
#subset <- subset[, sapply(subset, function(col) nlevels(col)) < 10]
#t1 <- tree(PRIMCAREYES ~. - PRIMCARE, data = subset, split = "gini")
#
#set.seed(40)
#t1cv <- cv.tree(t1, FUN = prune.misclass, data = subset)
#t1cv
```


## Revisiting the hypotheses {.unnumbered}
\doublespacing
Returning to the theoretical explanations for urgent care, it appears that while no one theory explains the totality of urgent care center usage, I observed much more evidence for hypotheses 1 and 3 than for 2---uprising given that the theory behind hypothesis two is the one most often cited in popular press stories about urgent care. There seems to be plenty of observable evidence to support hypothesis 3: there were a number of clusters which not only corresponded to the predicted trends for that hypothesis, but which directly stated that they use urgent care as their primary care provider.















