---
header-includes:
- \usepackage{graphicx,latexsym}
- \usepackage{amssymb,amsthm,amsmath}
- \usepackage{longtable,booktabs,setspace}
- \usepackage{tikz}
- \usepackage{setspace}
output: pdf_document
---
```{r rticle_rload, warning=FALSE, message=FALSE  ,echo=FALSE}
if(!require(devtools))
  install.packages("devtools", repos = "http://cran.rstudio.com")
if(!require(rticles)){
  library(devtools)
  devtools::install_github("ismayc/rticles")
  }
library(rticles)

setwd("~/Desktop/Thesis/Thesis")

```


# Appendix B {.unnumbered}
\onehalfspacing

#### K-modes / Why cluster? {.unnumbered}

In the supervised learning world, the data in question has an obvious response variable, which is tested against a null hypothesis based on theory. For this case, we do not know exactly what the outcome of interest is for those patients who went to urgent care, rather we are interested in the characteristics of who is choosing to go there. For such a data set, clustering methods allow us to examine the data in an *unsupervised* manner --- mainly in that we let the statistical software find the patterns rather than test for patterns at the start. 

#### Methodological Considerations {.unnumbered}

To perform the analysis on the data, four key methodological decisions were made by following either statistical convention or similar previous anlayses. All variables were re-coded as dummy variables, and the distances between these were standardized on a scale from 0 to 1, so as to prevent any skewdness which might result. Second, I chose to use the measure of distance known as the Jaccard method, which is specifically created to measure the distance between 0 to 1 scaled variables. It also has the unique feature of not including as significant pairs which both have 0 for a parameter. This is substantively important since though two visits may both have 0's for Private Insurance for example, the fact that they both don't have private insurance is not enough to consider them theoretically simliar by negation: one may be on medicare while the other may be uninsured. Third, for the actual groupings themselves, I have chosen the standard Ward's method, which attempts to minimize the variance within groups and thus maximizes the homogenaity. Fourth, in keeping with similar exploratory anlyses, I have limited the clusters to a theoretically interesting number while keeping a managable representation of reality. 

#### Determining the number of Clusters {.unnumbered}

Because urgent care centers have been greatly ignored by sociologists studying medical practices, there was little theoretical guidance in selecting a likely number of subgroups for the analysis. Similarly, because I am interested in understanding how an unsupervised analysis of patient data will reveal trends, it was particuarly important to the analysis that the number of clusters were both methematically acheivable and substantivly small enough for analysis. 

To accomplish this, I began with hierarchal clustering of the random samples chosen from the data. Using a mixed-methods tool to calculate the distance matrix between the various observations, I placed each point into an althgorithm which subsequently minimized the variance between clusters. Figure 1 shows the banner graph for the initial agglomerative cluster methods for the behavioral variables. 




```{r banr1, results = "asis", echo=FALSE}

rticles::label(path = "figures/banner.png", caption = "Sample Banner Agglomerative Clustering Results", 
      label = "ban1", type = "figure", scale = 0.35)
```

The white lines extending to the right represent clusters which differ from each other. After running the bottom up clustering method for a number of trials, the agglomerative coefficient was always between .73 and .8, indicating that as the height for which the clusters should stop combining. Again, Figure 1 demonstrates that at that height, around 8 clusters are have clearly seperated. 


#### OR {.unnumbered}

Because urgent care centers have been significantly overlooked by sociologists studying medical practices, there was little theoretical guidance in selecting a likely number of subgroups for the cluster analysis, often the first step in such examinations. Similarly, because I am interested in understanding how an unsupervised analysis of patient data will reveal trends, rather than imposing them onto the data, it was particularly important to the investigation that the number of clusters be both mathematically achievable and substantively small enough for analysis while still remaining unsupervised. The general rule of thumb for cluster analysis suggested by Mardia, Kent and Bibbby (1997) stipulates that the number of clusters k is approximately the square root of n / 2. However, in the case of our comparably small sample of urgent care visitors recorded in the NAMCS data set, such a rule would indicate at least 30 clusters—clearly not a workable number to make sense of patterns in urgent care patients and their use.

Instead, to estimate a k which would produce heterogeneous groups while still remaining theoretically enlightening, I began with non-guided hierarchal clustering of samples of 200 visits to urgent care centers randomly chosen from the data. Using the Gower methodology of finding the distance between dissimilar variable types, I calculated the distance matrix between the various observations for each sample. These distance measures were then used as the input to a hierarchal clustering algorithm which attempted to minimize the distance between observations within the same groups while maximizing the distance between them.  Figure 1 shows the ‘banner’ for the initial agglomerative cluster methods for the behavioral variables, and can be understood as a graphical representation of the points at which a cluster breaks away from the pack of observations. The white lines extending to the right represent clusters at their separation point, where larger red areas between white lines indicate stronger outside group variance.


It should be noted that though the analyses were performed in two waves, the clusters should be examined with each other in mind, and I have included some of the key demographic variables in with the behavioral variables to that end. [OH YEAH, THIS IS GOING IN THE APPENDIX]

*Figure 2. Color coded clusters.*

Figure 1 above shows the typical spread of the behavioral parameters’ clusters, and is useful for keeping the proportions of clusters in mind. 

