---
output: word_document
---
# Theoretical Foundations
  \setcounter{chapter}{2}
	\setcounter{section}{0}
  \doublespacing

```{r jess, echo=FALSE}
# apparently this is super jargony??
```

```{r from_matt, echo=FALSE}

# I think the most efficient way to connect them more clearly is to prioritize the parts of your theoretical/historical narrative that explain the actions of patients.

# Perhaps one thing to consider is moving the “Patient Choices” section beginning on page 11 to the beginning of the “Theoretical and Historical Foundations” chapter. If those arguments give the initial motivation for that chapter, they may also help you decide what needs more - and less - attention in the rest of the lit review.

# Your lit review moves the reader toward thinking that urgent care is replacing primary care and the strong relationships patients have had with doctors. I get the sense that your assumption is that urgent care patients would have otherwise gone to primary care physicians or emergency departments. But these new urgent care facilities may also be attracting patients who would have otherwise not gone to any doctor. Is there anything theoretical to add about that possibility?

# You discuss that patients have become increasingly demanding, attributing this to market-based theories about choice and beliefs about “shopping” for second options. Is there anything in the literature about other non-market explanations for patients becoming choosier? Speculating on whether things like WebMD or the popularity of media personalities like Dr Oz or Dr Phil have changed expectations about how people engage with medical institutions could be neat.

```


In this chapter, the quantitative exploratory analysis of urgent care centers that follows later will be located in the long-term primary care trends of patients seeking care within the American healthcare system. To comprehend the how and the why of the proliferation of urgent care centers, one must have some understanding of the complicated and tangled history of the medical profession in the US, and how institutional changes and epistomological shifts have led to the current state of affairs. The existing sociological research on the patient and practitioner, and the evolution of this relationship, are vitally important to beginning to understand how and at what level urgent care centers may be operating, since changes in the last half century have had drastic effects on this relationship.

I begin by tracing the development and subsequent decline of medical professional autonomy, paying particular attention to how structural changes dismantled the stronghold of political, economic and social power the profession claimed at its peak. I then review the current literature regarding the consequences of such structural changes on the doctor/patient relationship in the 21st century, taking up the traditional primary care physician's historical social importance. Combined, these changes lead us to the current ontologies of the American healthcare sphere where we will attempt to situate the role of urgent care centers and the patients who use them.

Through such a sociological lens, paying particular attention to how the literature understands the factors which influence the ways that patients access primary care. From this, I turn towards an exploration of primary care itself, and its importance and role in the larger healthcare field, its influence on literature around the patient/physician relationship, and the importance of the traditional primary care physician's historical social role. An examination of the profession shows a clear peak and subsequent decline in the power of the primary care's professional dominance in recent years and their primacy in the foctors which influence patient decisions, and the last portion of the background examines how structural changes may have helped to both dismantle the stronghold of political, economic and social power of the primary care physician, and what effect this may have on modern patients seeking primary care at urgent care centers.

## Patient Choices {.unnumbered} 

The last 30 years have seen a reconfiguration of the patient from passive recipient of care from their doctor to a critical consumer of health services [@barker_electronic_2008; @lupton_caveat_1991; @timmermans_continued_2010]. Theories of medical consumerism developed from economists studying the healthcare sector, and they begin with a similar hypothesis as the economic rational choice model, assuming that patients act as rational actors in the context of a medical encounter [@timmermans_continued_2010]. In other words, individuals act in a calculated manner to engage in self-improvement or health, and they are generally skeptical about expert knowledge [@lupton_consumerism_1997]. 

*Origins of the changing patient*
According to the literature, this trend began to express itself in the form of solicitation of second opinions and a sense of interchangeability of medical practitioners during the 1980’s, just as distrust of the industry reached its peak [@gray_trust_1997]. The idea that patients could shop around and compare services and prices was heavily popularized, and patients increasingly began to make autonomous decisions when selecting physicians [@hibbard_consumerism_1987; @lupton_caveat_1991].

Urgent care centers fit neatly into such a conceptualization of health services, and thus offer a key area of analysis in better understanding the developing roles of the consumer-patient within the larger healthcare industry. An important aspect of the research on the developing patient-consumers emphasizes the expansion of bargaining power on the part of the patient that came with the shift [@reeder_patient-client_1972]. A patient may now shop around the marketplace of health care, and that many are now choosing urgent care centers is undeniable given the industry’s rapid expansion. 

The shift towards urgent care centers undoubtedly plays some part in the transformations of the patient-practitioner relationship and the patient’s role in the medical care system discussed above, but so far these consequences remain largely unknown. Fortunately, sociologists have had a long-standing concern with such professional-client interactions, particularly within medicine [@freidson_sociology_1961; @mechanic_sociology_1966], and in order to better understand what effects the shifting norms around primary care physicians may have on patients, one can look to a large body of research on how institutional and organizational environments directly affect a patient’s experience in health care.

In 1963, Samuel Bloom published what would become one of the most cited sociological texts on the relationship between physicians and patients [@bloom_doctor_1965]. Even the title implies a power imbalance: the patient belongs to the doctor, whom distrbibutes his expert knowledge which is taken by the patient as truth on the basis of the profession's techno-cratic authority. 


Most of the literature locates the stability of this doctor-patient relationship in the fact that the physician acts as the patient’s agent, yet one can immediately recognize that urgent care centers may not be equipped to facilitate this relationship in the same way that the traditional primary care practice is seen to [@lupton_consumerism_1997]. The premise of urgency in such practices, the quickness with which patients are seen, and the targeted focus on acute problems all serve to create considerable doubt towards the ability of physicians within such an organizational context to fulfill the sociological role thought so important in previous literature.

Additionally, those who study health services have recognized that a patient’s medical history is a primary source of information regarding treatment, and primary care practitioners have been known to draw upon this as a valuable resource [@draper_primary-care_1975; @miller_primary_2010]. Urgent care centers however have no emphasis on maintaining patient-doctor ties and there is no reliable system to ensure a patient sees the same doctor even if they have been there before. 

### Why might patients not choose a traditioanl primary care practitioner? {.unnumbered}

Turning away from the fate of doctoring, it is important to acknowledge that the fate of primary care physicians has been reported to be hanging in the balance for quite some time. In 2005, the American College of Physicians released a report with the ominous title "The Impending Collapse of Primary Care Medicine and Its Implications for the State of the Nation's Health Care" [@ACP_impending_2006]. Yet while this may seem troublesome, a large body of recent sociological literature belongs to a growing number of scholars who are challenging the importance, and even relevance, of the traditional primary care physician in modern medicine. Those who have been observing developments in the patient-physician relationship over the past 30 years argue that the last quarter of the 20th century saw a dramatic reconfiguration of society, especially in regards to health services,  which has had profound effects on an individual’s relationship to the healthcare system. So while some were witnessing what was, for @starr_social_1982, “the social transformation of medicine,” and for many, “the end of the golden age of doctoring” [@mckinlay_end_2002]. 

## The Rise and Fall of the *Golden Age of Doctoring* {.unnumbered}

Since sociologists' initial interest in studying medicine and healthcare, the discipline was deeply intertwined with studies of the sociology of professionals and professionalism. In the U.S., Doctors were seen as the paragon of the *professional*: respected, organized, in control, and above all else, firmly established in their positions. Such traits led many to study what effects this had on medical care and how such professional dominance of medical practitioners shaped what services were provided. Many found that such dominance allowed the profession to block off areas of study entirely, or to ignore diseases they did not want to pursue. Such “modern doctors” worked within a “sovereign profession" (Starr 1982), serenely dispensing both medical care and authoritative judgment. Freidson comments that before the 1970’s, U.S. medicine “was at a historically unprecedented peak of prestige, prosperity and political and cultural influence—perhaps as autonomous as it is possible for a profession to be”(1988, p.384).

At the center of this prestige was the Physician. In the traditional Parsonian model, the Physician treated patients according to generalized technical standards of treatment (universalism), enacted a specific technical focus on medical care (specificity), treated patients without emotional investment (affective neutrality), and put the patient's welfare above their personal interests (the 'collective' orientation) (Parsons). Of course, even Parsons himself noted this was meant to function as a highly abstract view of the physician (1951, p.440), yet even in summary it bestows upon the doctor the scientific and moral high ground that was at one point firmly ingrained in the American psyche. 

Similar views of the doctor occur profusely throughout early to mid twentieth century literature on the profession, and is even visible today, though often as either a past to be reminisced on, or a goal to be returned to. A person's doctor, traditionally, was an extremely integrated, close, and important connection in a person's life. This was a relationship built, for some on trust, agency, and mutual understanding, while for others on power imbalance, technocratic authority and affective neutrality. When one thinks of the healthcare industry today, it is hard to call to mind the professional cohesion previously discussed. While doctors remain one of the more highly respected career choices in America, their prestige has certainly dropped since the 20th century [@heritage_problems_2006]. The term doctor is now applicable to a wide variety of sub-professions and specialties and the medical profession has become extremely diversified. While this may not have necessitated a drastic change in the societal role of the doctor, social scientists interested in this changing relationship have identified many possible sources of erosion.

#### The 21st century state has switched sides {.unnumbered}

```{r state_notes, echo=FALSE}
#you write that the state is “no longer supportive of health care.” This feels too strong given Obamacare, despite your suggestion that the ACA fits the “no longer supportive” narrative. You can still argue that it does, of course, but that claim needs more evidence. I expect most people think the ACA indicates greater governmental engagement and support with health care.
```

The government was an active and powerful ally in the rise of the medical profession during the 20th century (Alford, 1975). Political policies and government agencies served as buffers against what was 'legitimate' medicine and what was not, reaching the point where physicians achieved an almost monopolistic power over medical care. More than any other profession, the state served as legitimator and ally to medical professionals: state licensing for example, made it impossible for any other health care provider to legitimately practice medicine (Freidson, 1970a).

According to the literature, all of this began to unravel around the turn of the century, and in recent years the state has been transitioning towards an anti-leviathan New Right viewpoint, no longer supportive of health care (McKinlay & Marceau, 2002). This new approach avoids government intervention in economic and social life, consequently shifting the support of the state from the established medical organizations towards private interests. This shift in the primary allegiance of the state is reflected in many areas of social policy and has been clearly illustrated in recent attempts at health care reform, such as the Afforable Care Act.

Along with the government alliance shifts, big changes to Medicare and Medicaid legislation and the growth of third-party payers and for-profit medical service corporations created conditions which further removed the doctor from their traditional roles, eroding the political and cultural influence of the profession and threatening the cultural authority and technical autonomy of medicine (Starr 1982, Freidson 1988). 

#### Epidemiological transitions {.unnumbered}

Interestingly, one of the other most cited causes of the decline of primary care physicians, both in numbers and in prestige, is what is known as the epidemiological transition of developed countries (McKinley 2008). Medicine's rise around the middle of the 20th century coincided with the tail end of the infectious disease era. While there is academic disagreement over whether this decline was the result of medical organizations or an overall economic development, what is clear is that standards of living started to rise and Americans began to get sick *differently*. Rather than infectious diseases, which could strike anyone at any time, many argue that America has entered into a new epidemiological phase of chronic illness. According to one estimate, within 5 years (2020), an estimated 75 million Americans will have at least one chronic condition (Wu & Green, 2000).

While some could imagine that such chronic illnesses could secure the position of the primary care physician, McKinley argues that the new, chronically ill patient base is different in a fundamental way: "A well informed patient population can now obtain a proxy diagnosis on-line, or through health risk appraisals" (2008, p. 1483). Many patients know, or at least strongly suspect their ailment before ever considering a doctor and, according to McKinley, may even skip the initial diagnosis all together and immediately seek out a specialist.

#### Expansion of the playing field {.unnumbered}

Similarly to the retraction of government involvement, the last half of the 20th century saw a major expansion of health care actors. Physicians had the medical playing field almost entirely to themselves for most of the beginning of the 20th century, but this was no longer true as early as the 1970s. Non physician clinicians, unheard of 40 years ago, have increased dramatically in number, and what was once the sole field of the physician is now occupied by a range of actors (Starr, 1982; Cooper, Laud, & Dietrich, 1998). 

This is where urgent centers enter into the game most clearly: Late 20th century changes in the organization and financing of U.S. health care opened the door for the emergence of profit driven corporate care, and created a labor market for non-physician clinicians. For clinics which can staff a nurse practitioner for a fraction of the cost of a doctor, the choice is clear. Lower costs and customer satisfaction (two of the major tenets of the urgent care model) are widely accepted as 21st century health care priorities, and the result has been a dramatic surge in the competitive medical marketplace for health care. 

It should be noted that such changes may not have necessitated a loss of professional dominance. In fact, with the expansive growth on spending in the healthcare sector that the last 30 years have seen, it is plausible to imagine that doctors could have further cemented their professional medical authority. But most scholars have observed that the opposite has happened, and instead many view the past 30 years as the end of the authoritative medical professional. 

## From patient to client {.unnumbered}

These shifts in the professional dominance of the medical professional occurred parallel to changing norms surrounding the roles of both patients and practitioners, all of which directly affect a patient's acquisition and utilization of primary care. In 2006, the Millis report, commissioned by the American Medical Association, was asked to review the current status of physicians in the US. In the first section, they recalled the coveted physician of memory in America:

\singlespace

> "The general practitioner of revered memory knew his patients… and provided continuing care through the course of minor ailments and majors emergencies. His deficiencies… were partly offset by intimate knowledge of his patients, the support he gave them, and the trust and confidence his services engendered.”

\doublespacing

Intimate knowledge like that of a family practitioner who has known his patients for years is, according to both sociological and psychological research, the basis of a functional and stable doctor-patient relationship. A major sociological conceptualization of such relationships begins by speculating that the bureaucratization of modern social institutions has had drastic consequences on an individual's relationship to others and healthcare: as opportunities for close personal contacts diminish, problems which were originally handled in familial, social, and religious contexts are said to be transferred to ‘formal sustaining practitioners’ (Mechanic 1966). In modern societies, the prescribed structure of the 'doctor' and the 'patient' then provides a legitimization for the expression of intimacy and the request for help (1996). In other words, along with physical complaints and ailments, patients have coexisting and equally important socio-emotional needs, and sociologists of medicine have viewed addressing these needs as an essential task of the primary care physician.

But transformation is visible here as well: the “doctor” has become a “provider", the patient a "cliet". Juhar (2006) uses this "demise of the Physical Exam” as an illustration of these changing roles. He believes that patient testing technologies used in modern clinics are the newest in a long line of technologies and codified procedures which allow for *diagnosis at a distance*. Markel (2006, p551) agrees with Jauhar, and presents the stethoscope as the best symbol of medical practice, “… it embodies the essence of doctoring: using science and technology in concert with the human skill of listening to determine what ails the patient.” 

But he points out that, prior to the stethoscope, physicians would place an ear directly against a patient’s chest, a practice which inspired the inventor of the stethoscope only in that he found the act 'disgusting' and fashioned a listening technology which he could use to listen from a distance. Markel goes on to suggest that it was perhaps the stethoscope that began the process of physically separating physicians from their patients, and that with the rise of noninvasive imaging and testing and remote technologies, that distance will only continue to grow (2006). 


### Consequences for health care? {.unnumbered}

In light of such research, those who study at the intersection of health and social behaviors have begun to re-examine the importance of a close-tie relationship between doctor and patient in an attempt to respond to the consumer-patient model, developing a growing body of research which seeks to define the most effective components of patient-physician interactions and to reaffirm the place of the primary care physician in modern medicine. Common to most of these studies are the elements of ‘trust, compassion, communication, and clinical competence’ (Heritage and Maynard 2006; Phillips and Bazemore 2010, others at bottom). As an example, a 2002 study on clinical outcomes for low income women over the age of forty found that women who rated highest their doctor's ability to take care of all of their health care needs had 11 times the odds of ‘trusting their physician’ and 6 times the odds of finding their physicians ‘compassionate and communicative’, compared to those with the lowest level of comprehensiveness (O’Malley and Forrest 2002).

With the knowledge that close ties between physicians and patients have a direct impact on the perceived quality of care, the importance of examining the rise of urgent care centers become obvious. If patients are acting as rational consumers, which of them are choosing to receive their care outside of a primary care office, and what consequences does this have for their medical outcomes? The lack of research into the characteristics of urgent care patients make it difficult to answer such questions. Indeed, as of now, there has been no academic attempt to place the rapidly growing industry within the sociology of medicine, nor does there even exist an official definition of what constitutes an urgent care center. 

In the next section, three opposing hypotheses are developed regard how urgent care centers play a role in this new patient decision-making process. 






